Characterisation of transient resonances in extreme-mass-ratio-inspirals

### 2014-08-05 res_char.tex
adiabatic averaging

The radial and polar orbital periods in Mino time are given by
\begin{subequations}
\begin{align}
\Lambda_r = {} & 2\intd{r\sub{p}}{r\sub{a}}{\recip{\sqrt{V_r}}}{r} = \intd{-\pi}{\pi}{\diff{\lambda}{\psi}}{\psi}, \\
\Lambda_\theta = {} & 4\intd{\theta_-}{\pi/2}{\recip{\sqrt{V_\theta}}}{\theta} = \intd{-\pi}{\pi}{\diff{\lambda}{\chi}}{\chi}.
\end{align}
\end{subequations}
The orbital frequencies are thus
\begin{equation}
\Upsilon_r = \frac{2\pi}{\Lambda_r}, \quad \Upsilon_\theta = \frac{2\pi}{\Lambda_\theta}.
\end{equation}
The geodesic equations for coordinate time $t$ and azimuthal angle $\phi$ are just functions of $r$ and $\theta$, hence their evolutions can be expressed as Fourier series~\cite{Drasco2004}
\begin{subequations}
\begin{align}
\diff{t}{\lambda} = {} & \sum_{k_r,\,k_\theta}T_{k_r,\, k_\theta}\exp\left[-i\left(k_r\Upsilon_r + k_\theta\Upsilon_\theta\right)\lambda\right], \\
\diff{\phi}{\lambda} = {} & \sum_{k_r,\,k_\theta}\Phi_{k_r,\, k_\theta}\exp\left[-i\left(k_r\Upsilon_r + k_\theta\Upsilon_\theta\right)\lambda\right].
\label{eq:Mino-Fourier}
\end{align}
\end{subequations}
The $(0,\,0)$ coefficients in these series give the average secular rate of increase of these quantities. We define
\begin{equation}
\Gamma = T_{0,\,0}, \quad \Upsilon_\phi = \Phi_{0,\,0}
\end{equation}
to act as Mino time frequencies. We can now convert to coordinate time frequencies with
\begin{equation}
\Omega_r = \frac{\Upsilon_r}{\Gamma}, \quad \Omega_\theta = \frac{\Upsilon_\theta}{\Gamma}, \quad \Omega_\phi = \frac{\Upsilon_\phi}{\Gamma}.
\end{equation}

Transient resonances occur when the radial and poloidal motions are commensurate, when
\begin{equation}
\nu = \frac{\Upsilon_r}{\Upsilon_\theta} = \frac{\Omega_r}{\Omega_\theta} = \frac{n_\theta}{n_r}
\end{equation}
is the ratio of small integers. At this point any Fourier series like those in \eqnref{Mino-Fourier} goes from being an expansion in two frequencies to being an expansion in a single frequency~\cite{Bosley1992}.

For a general non-resonant orbit there is no fixed correlation between the radial and polar coordinates. After a sufficiently long time, the trajectory comes arbitrarily close to every point in the range of motion (with $r\sub{p} \leq r \leq r\sub{a}$ and $\theta_- \leq \theta \leq \pi - \theta_-$); on account of the orbital precession, the whole space is densely covered. This does not happen on resonance, as the radial and polar motions are locked together such that we can express one as a function of the other, and so the trajectory keeps cycling over the same path. The points visited are controlled by the relative phases of the $r$ and $\theta$ motions. To represent this, we use the $r$ phase at the $\theta$ turning point $\psi_{\theta_-} = \psi(\chi = 0)$. Varying $\psi_{\theta_-}$ across its full range allows every point in the range of motion to be reached. Hence averaging over all values of $\psi_{\theta_-}$ for a resonant orbit is equivalent to averaging over the $\psi$--$\chi$ $2$-torus for non-resonant orbits.

@@

For generic EMRIs, there are two timescales of interest: the \emph{fast} orbital motion, related to the fundamental frequencies $\sim1/\Omega$, and the \emph{slow} inspiral, related to the change of fundamental frequencies $\sim\Omega/\dot{\Omega}$. These, along with the resonance timescale, are discussed more in \secref{res-time}. The two-timescale nature of the problem makes it ideally suited to the method of osculating elements~\cite{Gair2011a}: on short timescales, we analyse the unperturbed system resulting in geodesic motion, and then the long-term evolution is described by a sequence of instantaneous geodesics.

@@

Given geodesic motion, forced by some external acceleration $a^\alpha$, so that
\begin{equation}
\difftwo{x^\alpha}{\tau} + \Gamma^\alpha_{\beta\gamma}\diff{x^\beta}{\tau}\diff{x^\gamma}{\tau} = a^\alpha,
\end{equation}
we can calculate the evolution of the osculating elements $\dot{I}^A$. The specific equations for motion in Kerr are derived by Gair et al \cite{Gair2011a}.

@@

Beyond geodesic motion in the Kerr spacetime, a test particle will follow an accelerated trajectory determined by \ref{eq:Mino-E-o-M}. As mentioned previously, this may be approximated by the adiabatic prescription~\cite{Hinderer2008} by dropping the forcing term $g_\alpha^{(1)}$ (and all higher order terms) and replacing $G_\alpha^{(1)}$ with its average over the $2$-torus parametrized by $q_r$ and $q_\theta$, $\langle G_\alpha^{(1)}\rangle_{q_r,\,q_\theta}$~\cite{Drasco2005}. This piece is purely dissipative and so signifies that flux is carried away from the system by gravitational waves.

In our parameterisation of the EMRI system, using $\{\psi,\chi\}$ rather than $\{q_r,q_\theta\}$, we must perform an appropriate average over the $2$-torus, given by
\begin{equation}
\label{eq:2torus-average}
\langle X\rangle_{\psi,\,\chi} = N \intd{0}{2\pi}{ \intd{0}{2\pi}{ \left(\partialdiff{\psi}{t}\right)^{-1} \left(\partialdiff{\chi}{t}\right)^{-1} \left(\diff{t}{\lambda}\right)^{-1} X }{\chi} }{\psi} .
\end{equation}
This form of averaging is successful at describing the leading order secular evolution of the trajectory, as is illustrated in \figref{good-traj}.

The effect of resonances must be completely contained within the pieces of the self-force left over after subtraction of the $2$-torus averaged pieces, $\hat{X}\equiv X-\langle X\rangle_{\psi,\,\chi}$ for some quantity $X$ related to the self-force. The adiabatic evolution, which contains only a single averaged forcing term, will therefore not demonstrate any resonance behaviours\footnote{Resonances still occur as the fundamental frequencies become commensurate, but as the system passes through each resonance, there is no additional effect on the evolution of the system.}. One method of studying the effect of resonances is then to compare an adiabatic evolution with an instantaneous evolution, using the full knowledge of the self-force. This may be a useful exercise to calibrate and verify adiabatic models that can be generated cheaply, however it cannot be used to distinguish between dephasing effects arising from resonances and those arising from conservative corrections to the phase variables, as these are both explicitly eliminated from the adiabatic prescription. We therefore take a different approach, comparing the instantaneous evolution with a quasi-adiabatic evolution, generated by averaging the full available self-force over the $2$-torus, rather than just averaging the term $G_\alpha^{(1)}$.

To be specific, we write \ref{eq:Mino-E-o-M} as
\begin{subequations}
\begin{align}
\label{eq:Mino-E-o-M2}
\diff{q_\alpha}{\lambda} = {} & \omega_\alpha(\boldsymbol{J}) + \eta \langle g_\alpha\rangle_{\psi,\,\chi}(\boldsymbol{J}) + \eta \hat{g}_\alpha(q_r,q_\theta,\boldsymbol{J}), \\
\diff{J_\alpha}{\lambda} = {} & \eta \langle G_\alpha\rangle_{\psi,\,\chi}(\boldsymbol{J}) + \eta \hat{G}_\alpha(q_r,q_\theta,\boldsymbol{J}),
\end{align}
\end{subequations}
where $g_\alpha$ and $G_\alpha$ encapsulate all our knowledge about the self-force and may contain contributions from sub-leading terms in the mass-ratio expansion. The quasi-adiabatic evolution is then obtained by dropping the terms proportional to $\hat{g}_\alpha$ and $\hat{G}_\alpha$. For ease of notation, we shall drop the \emph{quasi-} identifier; all subsequent references to an "adiabatic evolution" in fact refer to a "quasi-adiabatic evolution".

@@

The resonance jump depends sensitively on the relative phase of radial and poloidal motions. It is therefore instructive to study the resonance problem using an ensemble of systems differing only in the initial value of the radial phase $\psi$, which we do for our illustrative system parameters. Choosing random values for $\psi$ corresponds to random values of the orbital phase on resonance $\widehat{\kappa}_0$, which should map out the full range of resonance jumps for these particular orbital shape parameters, according to \eqref{eq:delta-I-a}.

Given our computed resonance jumps for all of the orbital parameters, we extract a single phase parameter $q$ for each system that specifies the magnitude of each jump. We expect such a phase parameter to be linearly offset from $\widehat{\kappa}_0$. The resonance jumps are sinusoidal with respect to $q$, as shown by the fits in \figref{resjump-vs-q}, lending support to \eqref{eq:delta-I-a} and giving us confidence in our numerical jump calculations. We can also compute the jumps in the fundamental frequencies for each system (shown as dashed lines in \figref{resjump-vs-q}); they are highly correlated to the magnitude of the jump in $E$, due to the dependence on $u_t$.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.46\textwidth]{Fig_jump_vs_q}
\caption{\label{fig:resjump-vs-q}The magnitude of the resonance jump for our illustrative system as a function of the extracted phase parameter $q$. The jump is expressed as a percentage of the adiabatic change in each parameter, given by \eqref{eq:res-jump-ratio}. The individual jumps as well as a sinusoidal fit are plotted for the three orbital shape parameters (solid lines) and the three orbital frequencies (dotted).}
\end{figure}

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{Fig_overlap_vs_resjump_ill}
\caption{\label{fig:overlap-vs-resjump-ill}The overlap \eqref{eq:overlap} between the full and adiabatic evolutions, maximised over the family of generated adiabatics, as a function of the resonance jumps in $E$ (left), $L_z$ (middle), and $Q$ (right). Each dot represents a system with our illustrative parameters and a random choice of the initial radial phase. The red line indicates the expected overlap assuming that the waveforms match exactly post-resonance, but have zero overlap pre-resonance.}
\end{figure*}

@@

High eccentricity EMRIs may also be possible, in which case transient resonances will play an important role in their detectability; systems that undergo a jump of a few percent dephase very rapidly from the adiabatic approximation. In such scenarios, adiabatic models can still be used, but there must be sufficient SNR in each non-resonant region that they can be identified as individual sources and then joined together as one long inspiral. Otherwise, if the phase difference on resonance can be predicted, \eqref{eq:delta-I-a} can be used to join together different adiabatic models.

>>>>

The radial and polar orbital periods in Mino time are given by
\begin{subequations}
\begin{align}
\Lambda_r = {} & 2\intd{r\sub{p}}{r\sub{a}}{\recip{\sqrt{V_r}}}{r} = \intd{-\pi}{\pi}{\diff{\lambda}{\psi}}{\psi}, \\
\Lambda_\theta = {} & 4\intd{\theta_-}{\pi/2}{\recip{\sqrt{V_\theta}}}{\theta} = \intd{-\pi}{\pi}{\diff{\lambda}{\chi}}{\chi}.
\end{align}
\end{subequations}
The orbital frequencies are thus
\begin{equation}
\Upsilon_r = \frac{2\pi}{\Lambda_r}, \quad \Upsilon_\theta = \frac{2\pi}{\Lambda_\theta}.
\end{equation}
The geodesic equations for coordinate time $t$ and azimuthal angle $\phi$ are just functions of $r$ and $\theta$, hence their evolutions can be expressed as Fourier series~\cite{Drasco2004}
\begin{subequations}
\begin{align}
\diff{t}{\lambda} = {} & \sum_{k_r,\,k_\theta}T_{k_r,\, k_\theta}\exp\left[-i\left(k_r\Upsilon_r + k_\theta\Upsilon_\theta\right)\lambda\right], \\
\diff{\phi}{\lambda} = {} & \sum_{k_r,\,k_\theta}\Phi_{k_r,\, k_\theta}\exp\left[-i\left(k_r\Upsilon_r + k_\theta\Upsilon_\theta\right)\lambda\right].
\label{eq:Mino-Fourier}
\end{align}
\end{subequations}
The $(0,\,0)$ coefficients in these series give the average secular rate of increase of these quantities. We define
\begin{equation}
\Gamma = T_{0,\,0}, \quad \Upsilon_\phi = \Phi_{0,\,0}
\end{equation}
to act as Mino-time frequencies. We can now convert to coordinate-time frequencies with
\begin{equation}
\Omega_r = \frac{\Upsilon_r}{\Gamma}, \quad \Omega_\theta = \frac{\Upsilon_\theta}{\Gamma}, \quad \Omega_\phi = \frac{\Upsilon_\phi}{\Gamma}.
\end{equation}

Transient resonances occur when the radial and poloidal motions are commensurate, when
\begin{equation}
\nu = \frac{\Upsilon_r}{\Upsilon_\theta} = \frac{\Omega_r}{\Omega_\theta} = \frac{n_\theta}{n_r}
\end{equation}
is the ratio of small integers. At this point, any Fourier series like those in \eqnref{Mino-Fourier} goes from being an expansion in two frequencies to being an expansion in a single frequency~\cite{Bosley1992}.

For a general non-resonant orbit there is no fixed correlation between the radial and polar coordinates. After a sufficiently long time, the trajectory comes arbitrarily close to every point in the range of motion (with $r\sub{p} \leq r \leq r\sub{a}$ and $\theta_- \leq \theta \leq \pi - \theta_-$); on account of the orbital precession, the whole space is densely covered. This does not happen on resonance, as the radial and polar motions are locked together such that we can express one as a function of the other, and so the trajectory keeps cycling over the same path. The points visited are controlled by the relative phases of the $r$ and $\theta$ motions. To represent this, we use the $r$ phase at the $\theta$ turning point $\psi_{\theta_-} = \psi(\chi = 0)$. Varying $\psi_{\theta_-}$ across its full range allows every point in the range of motion to be reached. Hence averaging over all values of $\psi_{\theta_-}$ for resonant orbits is equivalent to averaging over the $\psi$--$\chi$ $2$-torus for non-resonant orbits.

@@

For generic EMRIs, there are two timescales of interest: the \emph{fast} orbital motion, related to the fundamental frequencies $\sim1/\Omega$, and the \emph{slow} inspiral, related to the change in fundamental frequencies $\sim\Omega/\dot{\Omega}$, where an overdot denotes a derivative with respect to coordinate time $t$. These, along with the resonance timescale, are discussed more in \secref{res-time}. The two-timescale nature of the problem makes it ideally suited to the method of osculating elements~\cite{Gair2011a}: on short timescales, we analyse the unperturbed system resulting in geodesic motion, and then the long-term evolution is described by a sequence of instantaneous geodesics.

@@

Given geodesic motion in some background spacetime with Christoffel symbols $\Gamma^\alpha_{\beta\gamma}$, forced by some external acceleration $f^\alpha$, so that
\begin{equation}
\difftwo{x^\alpha}{\tau} + \Gamma^\alpha_{\beta\gamma}\diff{x^\beta}{\tau}\diff{x^\gamma}{\tau} = f^\alpha,
\end{equation}
we can calculate the evolution of the osculating elements $\dot{I}^A$. The specific equations for motion in Kerr are derived by Gair et al \cite{Gair2011a}.

@@

Beyond geodesic motion in the Kerr spacetime, a test particle follows an accelerated trajectory determined by \eqnref{Mino-E-o-M}. This may be approximated by the adiabatic prescription~\cite{Hinderer2008} by dropping the forcing term $g_\alpha^{(1)}$ (and all higher-order terms) and replacing $G_\alpha^{(1)}$ with its average over the $2$-torus parametrized by $q_r$ and $q_\theta$, $\langle G_\alpha^{(1)}\rangle_{q_r,\,q_\theta}$~\cite{Drasco2005}. This piece is purely dissipative and so signifies that flux is carried away from the system by GWs.

In our parameterisation of the EMRI system, using $\{\psi,\chi\}$ rather than $\{q_r,q_\theta\}$, we must perform appropriate averages over this new $2$-torus. The osculating elements integration scheme involves computing derivatives with respect to coordinate time, and we shall often be interested in the averaged form of such derivatives. Given a function $X(\psi,\chi)$, we first transform to Mino time where the motion is biperiodic
\begin{equation}
\langle \dot{X}\rangle_{\psi,\,\chi} \simeq \left.\left\langle \diff{X}{\lambda}\right\rangle_{\psi,\,\chi}\middle/\left\langle \diff{t}{\lambda}\right\rangle_{\psi,\,\chi}\right..
\end{equation}
It is then possible to calculate
\begin{widetext}\begin{equation}
\label{eq:2torus-average}
\left\langle \diff{X}{\lambda}\right\rangle_{\psi,\,\chi} = \frac1{\Lambda_r\Lambda_\theta} \intd{0}{2\pi}{ \intd{0}{2\pi}{ \left(\diff{\psi}{t}\right)^{-1} \left(\diff{\chi}{t}\right)^{-1} \left(\diff{t}{\lambda}\right)^{-1} \dot{X} }{\chi} }{\psi},
\end{equation}\end{widetext}
in terms of precomputed quantities. This form of averaging is successful at describing the leading order secular evolution of the trajectory, as illustrated in \figref{good-traj}.

The effect of resonances must be completely contained within the pieces of the self-force left after subtraction of the $2$-torus-averaged pieces, $\hat{X}\equiv X-\langle X\rangle_{\psi,\,\chi}$. The adiabatic evolution, which contains only a single $2$-torus-averaged forcing term, will therefore not be affected by the occurence of any transient resonances. Resonances still occur as the fundamental frequencies become commensurate, but as the system passes through each resonance, there is no additional effect on the evolution of the system.

One method of studying the effect of resonances is then to compare an adiabatic evolution with an instantaneous evolution, using the full knowledge of the self-force. This may be a useful exercise to calibrate and verify adiabatic models that can be generated cheaply; however it cannot be used to distinguish between dephasing effects arising from resonances and those arising from conservative corrections to the phase variables, as both are explicitly eliminated from the adiabatic prescription. We therefore take a different approach, comparing the instantaneous evolution with a quasi-adiabatic evolution, generated by averaging the full available self-force over the $2$-torus, rather than just averaging the term $G_\alpha^{(1)}$.

To be specific, we write \eqnref{Mino-E-o-M} as
\begin{subequations}
\begin{align}
\label{eq:Mino-E-o-M2}
\diff{q_\alpha}{\lambda} = {} & \omega_\alpha(\boldsymbol{J}) + \eta \langle g_\alpha\rangle_{\psi,\,\chi}(\boldsymbol{J}) + \eta \hat{g}_\alpha(q_r,q_\theta,\boldsymbol{J}), \\
\diff{J_\alpha}{\lambda} = {} & \eta \langle G_\alpha\rangle_{\psi,\,\chi}(\boldsymbol{J}) + \eta \hat{G}_\alpha(q_r,q_\theta,\boldsymbol{J}),
\end{align}
\end{subequations}
where $g_\alpha$ and $G_\alpha$ encapsulate all our knowledge about the self-force and may contain contributions from sub-leading terms in the mass-ratio expansion. The quasi-adiabatic evolution is then obtained by dropping the terms containing $\hat{g}_\alpha$ and $\hat{G}_\alpha$. For ease of notation, we shall drop the \emph{quasi-} identifier; all subsequent references to an ``adiabatic evolution'' in fact refer to a ``quasi-adiabatic evolution''.

@@

The resonance jump depends sensitively on the relative phase of radial and poloidal motions. It is therefore instructive to study the resonance problem using an ensemble of systems differing only in the initial value of the radial phase $\psi$, which we do for our illustrative system parameters. Choosing random values for $\psi$ corresponds to random values of the orbital phase on resonance $\widehat{\kappa}_0$, which should map out the full range of resonance jumps for these particular orbital shape parameters, according to \eqnref{delta-I-a}.

Given our computed resonance jumps for all of the orbital parameters, we extract a single phase parameter $q$ for each system that specifies the magnitude of each jump. We expect such a phase parameter to be linearly offset from $\widehat{\kappa}_0$. The resonance jumps are sinusoidal with respect to $q$, as shown by the fits in \figref{resjump-vs-q}, lending support to \eqnref{delta-I-a} and giving us confidence in our numerical jump calculations. We can also compute the jumps in the fundamental frequencies for each system (shown as dashed lines in \figref{resjump-vs-q}); they are highly correlated to the magnitude of the jump in $E$, due to the dependence on $u_t$.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.46\textwidth]{Fig_jump_vs_q}
\caption{\label{fig:resjump-vs-q}The magnitude of the resonance jump for our illustrative system as a function of the extracted phase parameter $q$. The jump is expressed as a percentage of the adiabatic change in each parameter, given by \eqnref{res-jump-ratio}. The individual jumps as well as a sinusoidal fit are plotted for the three orbital shape parameters (solid lines) and the three orbital frequencies (dotted).}
\end{figure}

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{Fig_overlap_vs_resjump_ill}
\caption{\label{fig:overlap-vs-resjump-ill}The overlap \eqnref{overlap} between the full and adiabatic evolutions, maximised over the family of generated adiabatics, as a function of the resonance jumps in $E$ (left), $L_z$ (middle), and $Q$ (right). Each dot represents a system with our illustrative parameters and a random choice of the initial radial phase. The red line indicates the expected overlap assuming that the waveforms match exactly post-resonance, but have zero overlap pre-resonance.}
\end{figure*}

@@

High eccentricity EMRIs may also be possible, in which case transient resonances will play an important role in their detectability; systems that undergo a jump of a few percent dephase very rapidly from the adiabatic approximation. In such scenarios, adiabatic models can still be used, but there must be sufficient SNR in each non-resonant region that they can be identified as individual sources and then joined together as one long inspiral. Otherwise, if the phase difference on resonance can be predicted, \eqnref{delta-I-a} can be used to join together different adiabatic models.
